---
title: "Multivariate predictive mean matching"
author: "Mingyang Cai"
date: ""
output: 
   html_document:
    toc: true
    toc_depth: 5
    toc_float: true
    number_sections: false
---
  
<style type="text/css">
  
body{ /* Normal  */
  font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 18px;
  color: DarkBlue;
}
h1 { /* Header 1 */
  font-size: 18px;
}
h2 { /* Header 2 */
  font-size: 18px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
  font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
  font-size: 14px;
}
</style>

---

Multivariate predictive mean matching is a generalized of univariate predictive mean, which can impute incomplete variables simultaneously.

# Packages used
The following packages are used.  
```{r message=FALSE, warning=FALSE}
library(mice)  
```

---

# Data generation
```{r}
set.seed(123)
B1 <- .5
B2 <- .5
X <- rnorm(1000)
XX <- X^2
e <- rnorm(1000, 0, 1)
Y <- B1 * X + B2 * XX + e
dat <- data.frame(x = X, xx = XX, y = Y)
# Impose 25 percent MCAR Missingness
dat[0 == rbinom(1000, 1, 1 - .25), 1:2] <- NA
```

---

# Imputation
```{r}
# Prepare data for imputation
blk <- list(c("x", "xx"), "y")
meth <- c("mpmm", "")
# Impute data
imp <- mice(dat, blocks = blk, method = meth, print = FALSE)
```

---

# Plot results
```{r}
# Pool result
pool(with(imp, lm(y ~ x + xx)))
plot(dat$x, dat$xx, col = mdc(1), xlab = "x", ylab = "xx")
cmp <- complete(imp)
points(cmp$x[is.na(dat$x)], cmp$xx[is.na(dat$x)], col = mdc(2))
```